---
title: "Liability threshold and time-to-event models"
author: "Advanced Topics in Genomics - Week 39"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
setwd("../liability_threshold_models/")
```

## Prerequisites
There are several requirements for completing this exercise.

1.  R studio with R version \>3.3.
2.  R package survival. For most users executing the `install.packages("survival")` ensures this.

```{r, echo = FALSE}
library(survival)
```

## Exercises

In these exercises, we will work on the concepts of the liability threshold model and time-to-event (survival) analysis.

We will:
 
 - Simulate genetic data under the liability threshold model
 
 - Estimate heritability from data with both genetic and environmental contributions
 
 - Explore the impact of ascertainment bias

 - Perform survival analysis to account for left and right censoring of data  

<br>

#### Part 1: Simulating data under the liability threshold model

The liability threshold model assumes that each individual has an underlying liability, which follows a normal distribution. Only individuals whose liability exceeds a certain threshold are considered cases.

```{r}

# Parameters
n_individuals <- 10000    # Number of individuals
n_snps <- 1000            # Number of SNPs that contribute to an individuals liability
maf <- 0.3                # Minor allele frequency
heritability <- 0.5       # Heritability of liability
threshold <- qnorm(0.95)  # Threshold 

# Simulate genotypes
set.seed(123)
genotypes <- matrix(rbinom(n_individuals * n_snps, 2, maf), nrow = n_individuals, ncol = n_snps)

# Simulate SNP effects
beta <- rnorm(n_snps, mean = 0, sd = sqrt(heritability / n_snps))

# Calculate liability
genetic_liability <- genotypes %*% beta

# Classify individuals as cases or controls based on threshold
phenotype <- ifelse(genetic_liability > threshold, 1, 0)

```

##### Q1: Make a histogram over the genetic liabilities. How many cases are there?

##### Q2: Look at the genotype matrix. What does 0, 1, 2 designate?

##### Q3: What does the threshold `threshold <- qnorm(0.95) ` indicate?

##### Q4: Look at line 56 in the code. How are the SNP effect sizes calculated and what assumption is it based on?

<br>

#### Part 2: Liability with an environmental component

We will now add an environmental component to the liability model to reflect the influence of non-genetic factors. This allows us to model a more realistic scenario where liability is influenced by both genetic and environmental factors.

```{r}
# Add environmental noise to the genetic liability
total_liability <- genetic_liability + rnorm(n_individuals, mean = 0, sd = sqrt(1 - heritability))

# Classify based on threshold
phenotype_env <- ifelse(total_liability > threshold, 1, 0)

```

##### Q5: Make a histogram over the liabilities.

##### Q6: How have we modeled environmental variance in the code above?

##### Q7: How does adding environmental noise affect the liability distribution? How does it influence the classification of individuals as cases or controls?

<br>

We can now estimate the heritability as the ratio of genetic variance to total variance.
```{r}

var_genetic <- var(genetic_liability)
var_total <- var(total_liability)

heritability_est <- var_genetic / var_total
heritability_est


```
##### Q8: Compare the estimated heritability with the true value used in the simulation. Why are these different?

<br>

#### Part 3: Ascertainment bias

Ascertainment bias occurs when cases and controls are sampled in a biased manner, often leading to over-representation of certain groups. This can significantly distort effect size estimates and heritability. 

In this exercise, we will simulate ascertainment bias by oversampling cases and investigate how this bias alters the data.

```{r}
# Sample individuals with bias (e.g., oversample cases)
set.seed(123)
ascertained_cases <- sample(which(phenotype_env == 1), 5000, replace = TRUE)
ascertained_controls <- sample(which(phenotype_env == 0), 5000, replace = TRUE)
ascertained_sample <- c(ascertained_cases, ascertained_controls)

# Visualize the ascertained data
liability_ascertained <- total_liability[ascertained_sample]
phenotype_ascertained <- phenotype_env[ascertained_sample]

hist(liability_ascertained, main = "Ascertainment Bias in Sampled Data", xlab = "Liability", xlim = c(-3, 3), breaks = 30, col = "lightcoral")
abline(v = threshold, col = "red", lwd = 2)


```

##### Q9: How does ascertainment bias alter the distribution of liability in the sample compared to before?

##### Q10: Reestimate the heritability. How has the estimate changed from before? Why?


<br>

#### Part 4: Accounting for left and right censoring
Survival analysis models are designed to handle time-to-event data and can account for right and left censoring. In this exercise, you will simulate time-to-event data and perform survival analysis using the Cox proportional hazards model.

We simulate both the time to event and censoring times (e.g., time when an individual's data is no longer available). The Cox model will help us analyze how genetic factors affect the risk of experiencing the event.

```{r}

set.seed(17)

# Identifying indices for cases and controls
case_indices <- which(phenotype_env == 1)
control_indices <- which(phenotype_env == 0)

# Simulate time-to-event using exponential distribution
# Time-to-event for cases
time_to_event_case <- rexp(length(case_indices), rate = 0.02)

# Time-to-event for controls
time_to_event_control <- rexp(length(control_indices), rate = 0.005) # We have modeled the events for cases to occur at a slightly higher rate, but this is arbitrary

# Create a time_to_event vector in the same order as the phenotype vector
time_to_event <- numeric(length(phenotype_env))  # Initialize an empty vector
time_to_event[case_indices] <- time_to_event_case  # Assign event times to cases
time_to_event[control_indices] <- time_to_event_control  # Assign follow-up times to controls

# Create event status: 1 for cases (event happened), 0 for controls (censored)
event_status <- phenotype_env

# Add the first SNP as a covariate for the CoxPH model (for demonstration)
snp1 <- genotypes[, 1]

# Fitting the model
cox_model <- coxph(Surv(time_to_event, event_status) ~ snp1)

# Summarize the CoxPH model
summary(cox_model)

```
##### Q11: How does the estimated hazard ratio from the Cox model help us understand the effect of SNP1 on disease onset?

```{r}
lambda_death_case <- 0.1  # Slightly increase case event rate
lambda_death_control <- 0.06  # Slightly reduce control event rate

# Simulate time to event for cases and controls
time_to_death_case <- rexp(sum(phenotype_env == 1), rate = lambda_death_case)
time_to_death_control <- rexp(sum(phenotype_env == 0), rate = lambda_death_control)

# Combine time to event into a single vector
time_to_event <- numeric(length(phenotype_env))  # Initialize empty vector
time_to_event[phenotype_env == 1] <- time_to_death_case  # Assign event times to cases
time_to_event[phenotype_env == 0] <- time_to_death_control  # Assign event times to controls

# Simulate censoring: Assume some individuals are censored
event_status <- rbinom(n_individuals, 1, 0.7)  # 70% of individuals experience the event

# Kaplan-Meier curve stratified by phenotype
km_fit <- survfit(Surv(time_to_event, event_status) ~ phenotype_env)

# Plot the KM curve
plot(km_fit, col = c("blue", "red"), lty = 1:2, xlab = "Time", ylab = "Survival Probability",
     main = "Kaplan-Meier Curve")
legend("topright", legend = c("Control", "Case"), col = c("blue", "red"), lty = 1:2)

```

##### Q12: How does the survival curve differ between cases and controls?
